\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-11-30 chapters/covert_align.tex
Accuracy figure
Preface, ironed out overhangs

The decoding methods described in \cref{sec:stbf-struct} and
\cref{sec:bttda} are general attempts at increasing \ac{erp} decoding accuracy.
Given the problem statement of this work, we are also interested in targeted
methods that would specifically advance the development of gaze-independent
visual \acp{bci} (see \cref{sec:gaze-independence/sota}).
Methods of directly enhancing gaze-independent decoding through optimizing
decoder design have not been widely studied (see
\cref{sec:gaze-independence/sota/decoding}).
We assume that advances in gaze-independent decoding can be made if properties
of covert \ac{vsa} that affect performance can be identified and accounted for.
The latency estimation and alignment approach proposed in \cref{sec:wcble}
was specifically designed for this purpose.

@@

Participants signed an informed consent form and were seated at a distance of
60 cm from a CRT-emulating monitor (VPixx
Technologies, Canada) operating at a refresh rate of 120 Hz, displaying 6
circular white targets with a diameter of 4.15° visual angle and laid out on a hexagon
with a radius of 12.28° of visual angle centered on the midpoint of the screen,
conforming to the interface proposed by \textcite{Treder2010}
(\cref{fig:covert-align/interface/interface}).
A hexagonal layout interface with an empty center and a low number of targets
counteracts target crowding, and as long as the subject’s gaze is within the hexagon of
targets, no other target can be between the subject’s gaze and a covertly
attended target.
Targets are full-contrast white and were intensified by scaling them to a
larger size (5.60° of visual angle, see \cref{fig:covert-align/interface/intensification})
instead of changing the contrast to avoid Troxler-fading\footnote{The optical illusion of disappearing unchanging stimuli
experienced when visually fixating~\cite{Troxler1804}.}~\cite{Treder2010} in the
peripheral visual field.
Stimuli were presented using PsychoPy (version 2023.1.3)~\cite{Peirce2019}.

@@

The experiment started with five non-recorded practice stimulation blocks, one for
each of the five \ac{vsa} conditions.
During these practice blocks, the participant received feedback about their gaze
position and counting accuracy.
Counting the instructions and the participant's response to the
input prompts, a block lasted about 30 seconds. In sum, the experiment featured approximately 45 minutes of
stimulation time.
After blocks 14 and 28, the participant was allowed to take a longer break.
Including these longer breaks, the experiment lasted approximately one hour.

@@

\Cref{fig:covert-align/block-eval} shows the cross-validated \ac{bci} selection
accuracy on the BNCI2014-009 and the CVSA-ERP datasets for all investigated
decoders.
Accuracy was determined by, for each block, selecting the character with the
highest (stage-two if applicable) classifier score and comparing it to the cued target.
Significance was calculated using one-sided Wilcoxon signed-rank tests ($p=0.05$)
corrected for FDR over decoders.
For this evaluation, no epochs were rejected to keep the trial-based structure
of \ac{bci} operation intact.
For all datasets and \ac{vsa} conditions, \ac{cble} scores approximately on par with \ac{tlda}.
Yet, \ac{wcble} yields an improved decoding accuracy for covert \ac{vsa} in both
datasets, which is greatest for smaller numbers of repetitions and
decreases as the number of repetitions increases.
This covert \ac{vsa} accuracy increase over \ac{tlda} is significant in the
BNCI2014-009 dataset for 1 and 3 repetitions, and in CVSA-ERP for 1-5 and 10 repetitions.
Furthermore, while we reported a relative decrease in single-trial \ac{rocauc} for
\ac{wcble} in overt \ac{vsa}, this does not seem to result in a consistent decrease in target
selection accuracy.
No significant increase of \ac{wcble} over other methods was found in split \ac{vsa}.
While Riemannian methods are significantly outperformed by \ac{tlda}, \ac{cble}, and \ac{wcble}
in the BNCI2014-009 dataset, they perform approximately on par with \ac{tlda} and
\ac{wcble} in CVSA-ERP.
\begin{figure}[t]
		\label{fig:covert-align/block-eval}%
		\includegraphics[width=\linewidth]{figures/covert_align/figure4b.pdf}
  \caption[Selection accuracy.]{%
		Cross-validated target selection accuracy for
    all decoders plotted as a function of the number of test repetitions in
    different VSA conditions. Significance was determined using one-sided
    (\ac{wcble} $>$ other) Wilcoxon signed-rank tests using False Discovery Rate correction
    ($*= p<0.05$, $**=p<0.01$, $***=p<0.001$). \ac{wcble} generally achieves the highest
    covert \ac{vsa} target selection accuracy.}
\end{figure}

>>>>

The decoding methods described in \cref{sec:stbf-struct} and
\cref{sec:bttda} are general attempts at increasing \ac{erp} decoding accuracy.
Given the problem statement of this work, we are also interested in targeted
methods that would specifically advance the development of gaze-independent
visual \acp{bci} (see \cref{sec:gaze-independence/sota}).
Methods that directly enhance gaze-in\-de\-pen\-dent decoding through optimizing
decoder design have not been widely studied (see
\cref{sec:gaze-independence/sota/decoding}).
We assume that advances in gaze-independent decoding can be made if properties
of covert \ac{vsa} that affect performance can be identified and accounted for.
The latency estimation and alignment approach proposed in \cref{sec:wcble}
was specifically designed for this purpose.

@@

Participants signed an informed consent form and were seated at a distance of
60 cm from a CRT-emulating monitor (VPixx
Technologies, Ca\-na\-da) operating at a refresh rate of 120 Hz, displaying 6
circular white targets with a diameter of 4.15° visual angle and laid out on a hexagon
with a radius of 12.28° of visual angle centered on the midpoint of the screen,
conforming to the interface proposed by \textcite{Treder2010}
(\cref{fig:covert-align/interface/interface}).
A hexagonal layout interface with an empty center and a low number of targets
counteracts target crowding, and as long as the subject’s gaze is within the hexagon of
targets, no other target can be between the subject’s gaze and a covertly
attended target.
Targets are full-contrast white and were intensified by scaling them to a
larger size (5.60° of visual angle, see \cref{fig:covert-align/interface/intensification})
instead of changing the contrast to avoid Troxler-fading\footnote{The optical illusion of disappearing unchanging stimuli
experienced when visually fixating~\cite{Troxler1804}.}~\cite{Treder2010} in the
peripheral visual field.
Stimuli were presented using PsychoPy (version 2023.1.3)~\cite{Peirce2019}.

@@

The experiment started with a sequence  of five non-recorded practice stimulation blocks, one for
each of the five \ac{vsa} conditions.
During these practice blocks, the participant received feedback about their gaze
position and counting accuracy.
Counting the instructions and the participant's response to the
input prompts, a block lasted about 30 seconds. In sum, the experiment featured approximately 45 minutes of
stimulation time.
After blocks 14 and 28, the participant was allowed to take a longer break.
Including these longer breaks, the experiment lasted approximately one hour.

@@

\Cref{fig:covert-align/block-eval} shows the cross-validated \ac{bci} selection
accuracy on the BNCI2014-009 and the CVSA-ERP datasets for all investigated
decoders.
Accuracy was determined by, for each block, selecting the character with the
highest (stage-two if applicable) classifier score and comparing it to the cued target.
Significance was calculated using one-sided Wilcoxon signed-rank tests ($p=0.05$)
corrected for FDR over decoders.
For this evaluation, no epochs were rejected to keep the trial-based structure
of \ac{bci} operation intact.
For all datasets and \ac{vsa} conditions, \ac{cble} scores approximately on par with \ac{tlda}.
Yet, \ac{wcble} yields an improved decoding accuracy for covert \ac{vsa} in both
datasets, which is greatest for smaller numbers of repetitions and
decreases as the number of repetitions increases.
This covert \ac{vsa} accuracy increase over \ac{tlda} is significant in the
BNCI2014-009 dataset for 1 and 3 repetitions, and in CVSA-ERP for 1-5 and 10 repetitions.
Furthermore, while we reported a relative decrease in single-trial \ac{rocauc} for
\ac{wcble} in overt \ac{vsa}, this does not seem to result in a consistent decrease in target
selection accuracy.
No significant increase of \ac{wcble} over other methods was found in split \ac{vsa}.
While Riemannian methods are significantly outperformed by \ac{tlda}, \ac{cble}, and \ac{wcble}
in the BNCI2014-009 dataset, they perform approximately on par with \ac{tlda} and
\ac{wcble} in CVSA-ERP.
\begin{figure}[t]
		\label{fig:covert-align/block-eval}%
		%\includegraphics[width=\linewidth]{figures/covert_align/figure4b.pdf}
    \input{figures/covert_align/block_accuracy.pgf}
  \caption[Selection accuracy.]{%
		Cross-validated target selection accuracy for
    all decoders plotted as a function of the number of test repetitions in
    different VSA conditions. \ac{wcble} generally achieves the highest
    covert \ac{vsa} target selection accuracy.}
\end{figure}

### 2024-11-30 chapters/cv.tex
Started processing comments MDV
Remove at
Remove at and .tex in include
Preface
Preface, ironed out overhangs

\subsection*{Awards \& funding}
\begin{cvlist}
	\cvitem{Clinical Trial Sponsorship Program}{Zeto, 2024 \\
		Our research project titled ``A gaze-independent Visual Brain-Computer
		Interface for use by patients with limited or no eye control'' won a spot in
		the Zeto Clinical Trial Sponsorship Program, granting us the resources to conduct research
		with the Zeto EEG headset.
	}
	\cvitem{Student BCI competition 4th place worldwide}{NeuroTechX, 2022 \\
		Our NeuroTechLeuven BrainBrowsR project, a plug-and-play BCI system that lets you control
		social media applications  trough SSVEP-BCI was awarded 90 out of 100 points by
		the jury of the NeuroTechX international student club competition.
	}
	\cvitem{Global PhD Partnership Grant}{KU Leuven and University of Lille, 2021--2024}
\end{cvlist}

\subsection*{Conferences and presentations}
\begin{cvlist}
	\cvitem{CORTICO Days}{CORTICO, Nancy, 2024 \\
		Poster presentation on covert visual attention BCIs for patients with
		oculomotor impairment.}
	\cvitem{10th International BCI Meeting}{BCI Society, Brussels, 2023\\
		Poster presentation on classifier-based latency estimation for
		gaze-independent BCIs.}
	\cvitem{Leuven.AI Scientific Workshop 2022}{Leuven.AI, Leuven, 2022\\
		Poster presentation on (multi-)Kronecker-structured linear discriminant
		analysis for low sample size event-related potential classification.}
	\cvitem{CORTICO Days}{CORTICO, Grenoble, 2022\\
		Presentation on Kronecker-structured LCMV-beamforming for event-related potential
		classification.}
	\cvitem{CORTICO Days}{CORTICO, 2021\\
		Presentation on a multi-component approach
		to spatiotemporal beamforming decoding of event-related potentials.}
	\cvitem{BCI \& Neurotechnology Spring School 2024}{g.tec, 2024}
	\cvitem{Closed Loop Neurotechnologies Autumn School}{NeurotechEU, Lille, 2023}
	\cvitem{CORTICO Days}{CORTICO, Paris, 2023}
	\cvitem{BCI \& Neurotechnology Spring School 2023}{g.tec, 2023}
	\cvitem{BCI \& Neurotechnology Masterclass Belgium}{g.tec, 2022}
	\cvitem{BCI \& Neurotechnology Spring School 2022}{g.tec, 2022}
\end{cvlist}

@@

\subsubsection{Master theses and internships supervised}
\begin{cvlist}
	\cvitem{Eye-tracker and ERP data fusion for gaze-independent visual
		BCI}{\ \\ Juliette  Meunier, University of Lille, 2024}
	\cvitem{LDA in	the combined
		space-time-frequency domain for BCI decoding}{\ \\Lunkyadi Sucipto, KU Leuven,
		2023--2023}
	\cvitem{Tackling the Midas Touch problem in eye tracking with a
		BCI}{\ \\Reniflal Ebenezer Sundaralal, KU Leuven, 2023--2023}
	\cvitem{An EEG connectivity analysis of Alzheimer’s Disease
		and Frontotemporal Dementia}{Zoe Barinaga, KU Leuven, 2023--2024}
	\cvitem{Single-trial ERP latencies as a predictor for the mode
		of visual attention}{\ \\Yildiz Dilara Parry, KU Leuven, 2023}
	\cvitem{A hybrid P300-gaze BCI alternative for navigating virtual
		spaces}{\ \\Gijs Claes, KU Leuven, 2021--2022}
\end{cvlist}

@@

\subsection*{Languages}
\begin{cvlist}
	\cvitem{Dutch}{native}
	\cvitem{English}{fully proficient}
	\cvitem{French}{advanced}
	\cvitem{German}{intermediate}
\end{cvlist}

>>>>

\subsection*{Awards \& funding}
\begin{cvlist}
	\cvitem{Clinical Trial Sponsorship Program}{Zeto, 2024 \\
		Our research project titled ``A gaze-independent Visual Brain-Com\-pu\-ter
		Interface for use by patients with limited or no eye control'' won a spot in
		the Zeto Clinical Trial Sponsorship Program, granting us the resources to conduct research
		with the Zeto EEG headset.
	}
	\cvitem{Student BCI competition 4th place worldwide}{NeuroTechX, 2022 \\
		Our NeuroTechLeuven BrainBrowsR project, a plug-and-play BCI system that lets you control
		social media applications  trough SSVEP-BCI was awarded 90 out of 100 points by
		the jury of the NeuroTechX international student club competition.
	}
	\cvitem{Global PhD Partnership Grant}{KU Leuven and University of Lille, 2021--2024}
\end{cvlist}

\subsection*{Conferences and presentations}
\subsubsection*{Presented}
\begin{cvlist}
	\cvitem{CORTICO Days}{CORTICO, Nancy, 2024 \\
		Poster presentation on covert visual attention BCIs for patients with
		oculomotor impairment.}
	\cvitem{10th International BCI Meeting}{BCI Society, Brussels, 2023\\
		Poster presentation on classifier-based latency estimation for
		gaze-in\-de\-pen\-dent BCIs.}
	\cvitem{Leuven.AI Scientific Workshop 2022}{Leuven.AI, Leuven, 2022\\
		Poster presentation on (multi-)Kronecker-structured linear discriminant
		analysis for low sample size event-related potential classification.}
	\cvitem{CORTICO Days}{CORTICO, Grenoble, 2022\\
		Presentation on Kronecker-structured LCMV-beamforming for event-related potential
		classification.}
	\cvitem{CORTICO Days}{CORTICO, 2021\\
		Presentation on a multi-component approach
		to spatiotemporal beamforming decoding of event-related potentials.}
\end{cvlist}

\subsubsection*{Attended}
\begin{cvlist}
	\cvitem{BCI \& Neurotechnology Spring School 2024}{g.tec, 2024}
	\cvitem{Closed Loop Neurotechnologies Autumn School}{NeurotechEU, Lille, 2023}
	\cvitem{CORTICO Days}{CORTICO, Paris, 2023}
	\cvitem{BCI \& Neurotechnology Spring School 2023}{g.tec, 2023}
	\cvitem{BCI \& Neurotechnology Masterclass Belgium}{g.tec, 2022}
	\cvitem{BCI \& Neurotechnology Spring School 2022}{g.tec, 2022}
\end{cvlist}

@@

\subsubsection*{Master theses and internships}
\begin{cvlist}
	\cvitem{Eye-tracker and ERP data fusion for gaze-independent visual
		BCI}{\ \\ Juliette  Meunier, University of Lille, 2024}
	\cvitem{LDA in	the combined
		space-time-frequency domain for BCI decoding}{\ \\Lunkyadi Sucipto, KU Leuven,
		2023--2023}
	\cvitem{Tackling the Midas Touch problem in eye tracking with a
		BCI}{\ \\Reniflal Ebenezer Sundaralal, KU Leuven, 2023--2023}
	\cvitem{An EEG connectivity analysis of Alzheimer’s Disease
		and Frontotemporal De\-men\-tia}{Zoe Barinaga, KU Leuven, 2023--2024}
	\cvitem{Single-trial ERP latencies as a predictor for the mode
		of visual attention}{\ \\Yildiz Dilara Parry, KU Leuven, 2023}
	\cvitem{A hybrid P300-gaze BCI alternative for navigating virtual
		spaces}{\ \\Gijs Claes, KU Leuven, 2021--2022}
\end{cvlist}

@@

%\subsection*{Languages}
%\begin{cvlist}
%	\cvitem{Dutch}{native}
%	\cvitem{English}{fully proficient}
%	\cvitem{French}{advanced}
%	\cvitem{German}{intermediate}
%\end{cvlist}

### 2024-11-30 chapters/gaze_independence.tex
Preface, ironed out overhangs

\subsection{Gaze-independent modalities}
Gaze-independent \ac{erp}-based \acp{bci}~\cite{Riccio2012, Aloise2012} can be realized in three
ways.
Firstly, active \ac{bci} communication paradigms relying on endogenous activation from the user
do not rely on sensory stimulation.
Examples of this are imagined movement or imagined speech paradigms.
These paradigms can yield very high information transfer
rates~\cite{Willett2021,Metzger2023}, both due to their intuitiveness and the
complexity that can be captured in commands, but they often do so only when paired
with invasive recording.
Non-visual reactive paradigms that use, for example, auditory~\cite{Halder2010} and
somatosensory~\cite{Brouwer2010} stimulation do not rely on gaze redirection
but also result in lower \acp{itr} compared to visual paradigms.
A survey by \textcite{Riccio2012} from 2012 reported that \acp{itr} ranged between
1 and 10 bits/min for auditory paradigms and 0.5 and 5 bits/min for tactile~\cite{Riccio2012}.

@@

We make a distinction here between spatially organized interfaces, where
multiple targets are displayed at the same time at different spatial locations,
and temporally organized interfaces, like \ac{rsvp}, where targets or small
sets of targets (e.g. 3) are shown consecutively.
The Hex-o-Spell, GeoSpell, and GIBS Block Speller are spatially organized, while
variants of the \ac{rsvp} paradigm are temporally organized.
\textcite{Lin2018} follow the philosophy that spatially organized interfaces
have historically not performed well in the presence of gaze impairment.
However, we argue that spatially organized interfaces generally have a much higher
\ac{itr} in healthy controls.
For instance, a regular matrix-based speller can achieve an \ac{itr} of up to
30 bits/min.
With the right set of adaptations, spatial attention could potentially also
bring added value to individuals with \ac{sspgi}.
While spatial interfaces are indeed more gaze-dependent than temporal
interfaces, they also provide an extra channel that can be used to transfer
information, i.e., the location of the stimulus.

@@

The previous two methods did not yet yield specific results in gaze-independent
decoding.
The following approach is targeted at improving decoding performance in
gaze-independent settings by accounting for a known property of the covert
attention \ac{erp} response.

@@

Existing latency estimation methods are either not applicable to the
classification problem of labeling unseen data, or are not robust enough to
deal with the low \ac{snr} of the \ac{erp}.
\Ac{cble}~\cite{Mowla2017} is a technique that can leverage
\ac{erp} latency estimation in a decoding setting, but our results show that it
yields no improvement in gaze-independent settings.
We improved upon this technique and extended it to a probabilistic, iterative
method named \ac{wcble}.
This approach has been published by~\textcite{VanDenKerchove2024} and is presented
and evaluated on simulated data in \cref{sec:wcble}.
We have also collected datasets to evaluate this approach on real \ac{erp} data
in gaze-independent settings.

@@

The hypothesis was that gaze-independent decoders, such as \ac{wcble}~\cite{VanDenKerchove2024},
would improve performance in covert and free \ac{vsa} conditions compared to standard
decoders like \ac{tlda}.
We aimed to determine if calibration in an overt \ac{vsa} setting could also enhance
performance in free and covert conditions, leveraging residual eye motor control during
calibration to improve overall \ac{bci} accuracy for gaze-impaired users.
The results of this study are presented in \cref{sec:patients}.

>>>>

\subsection{Gaze-independent modalities}
Gaze-independent \ac{erp}-based \acp{bci}~\cite{Riccio2012, Aloise2012} can be realized in three
ways.
Firstly, active \ac{bci} communication paradigms relying on endogenous activation from the user
do not rely on sensory stimulation.
Examples of this are imagined movement or imagined speech paradigms.
Ac\-tive paradigms can yield very high information transfer
rates~\cite{Willett2021,Metzger2023}, both due to their intuitiveness and the
complexity that can be captured in commands, but they often do so only when paired
with invasive recording.
Non-visual reactive paradigms that use, for example, auditory~\cite{Halder2010} and
somatosensory~\cite{Brouwer2010} stimulation do not rely on gaze redirection
but also result in lower \acp{itr} compared to visual paradigms.
A survey by \textcite{Riccio2012} from 2012 reported that \acp{itr} ranged between
1 and 10 bits/min for auditory paradigms and 0.5 and 5 bits/min for tactile~\cite{Riccio2012}.

@@

We make a distinction here between spatially organized interfaces, where
multiple targets are displayed at the same time at different spatial locations,
and temporally organized interfaces, like \ac{rsvp}, where targets or small
sets of targets (e.g. 3) are shown consecutively.
The Hex-o-Spell, GeoSpell, and GIBS Block Speller are spatially organized, while
variants of the \ac{rsvp} pa\-ra\-digm are temporally organized.
\textcite{Lin2018} follow the philosophy that spatially organized interfaces
have historically not performed well in the presence of gaze impairment.
However, we argue that spatially organized interfaces generally have a much higher
\ac{itr} in healthy controls.
For instance, a regular matrix-based speller can achieve an \ac{itr} of up to
30 bits/min.
With the right set of adaptations, spatial attention could potentially also
bring added value to individuals with \ac{sspgi}.
While spatial interfaces are indeed more gaze-dependent than temporal
interfaces, they also provide an extra channel that can be used to transfer
information, i.e., the location of the stimulus.

@@

The previous two methods did not yet yield specific results in
gaze-in\-de\-pen\-dent
decoding.
The following approach is targeted at improving decoding performance in
gaze-independent settings by accounting for a known property of the covert
attention \ac{erp} response.

@@

Existing latency estimation methods are either not applicable to the
classification problem of labeling unseen data, or are not robust enough to
deal with the low \ac{snr} of the \ac{erp}.
\Ac{cble} as described by \textcite{Mowla2017} is a technique that can leverage
\ac{erp} latency estimation in a decoding setting, but our results show that it
yields no improvement in gaze-independent settings.
We improved upon this technique and extended it to a probabilistic and iterative
me\-thod, named \ac{wcble}.
This approach has been published by~\textcite{VanDenKerchove2024} and is presented
and evaluated on simulated data in \cref{sec:wcble}.
We have also collected datasets to evaluate this approach on real \ac{erp} data
in gaze-independent settings.

@@

The hypothesis was that gaze-independent decoders, such as \ac{wcble},
would improve performance in covert and free \ac{vsa} conditions compared to standard
decoders like \ac{tlda}.
We aimed to determine if calibration in an overt \ac{vsa} setting could also enhance
performance in free and covert conditions, leveraging residual eye motor control during
calibration to improve overall \ac{bci} accuracy for gaze-im\-pair\-ed users.
The results of this study are presented in \cref{sec:patients}.

### 2024-11-30 chapters/patients.tex
Accuracy figure
Preface, ironed out overhangs

\Acp{bci} for communication assistive technology~\cite{Millan2010}
target individuals with \acf{sspi}~\cite{Peters2022}.
Visual \acp{bci}, which rely on the interpretation of visual stimuli by the user,
offer several advantages in this context.
They can work with non-invasive recording technology and can use rapid
stimulation.
This makes them well-suited for real-time communication tasks.

@@

Several studies with visual oddball \acp{bci} show that performance drops when not fixating the intended
target~\cite{Brunner2010, Treder2010, RonAngevin2019}, necessitating
gaze-independent solutions.
These studies build on the assumption that \ac{bci} users with \ac{sspgi}
would feel comfortable operating an interface in pure covert \ac{vsa} with
central fixation.
One could argue that a \ac{bci} that is only verified to work when central
fixation is maintained could also be considered gaze-dependent.
This does not account for the residual eye motor capabilities of most people
with \ac{sspi}, the (dis)comfort they experience while
performing gaze fixation and other confounding factors resulting from their eye
motility.

@@

\textcite{Lesenfants2014} tested a gaze-independent \ac{ssvep} \ac{bci} in six
participants with \ac{lis} yet only exceeded chance level accuracy in two.
More recently, \textcite{Peters2020} performed a trial with two
participants with late-stage \ac{als} and visual impairment.
Their \ac{ssvep} interface was not optimized for gaze-independence, but the
system showed high accuracy, outperforming an eye tracking alternative.
It would be of interest to verify if such results can be replicated with
participants with other conditions, and with a visual oddball \acp{bci}.

@@

\section{Methods}
\subsection{Recruitment}
Participants were recruited across the Neuromuscular Reference center at
University Hospital Leuven (Leuven, Belgium), TRAINM Neuro Rehab Clinics
(Antwerp, Belgium), the Neurorehabilitation Unit at University Hospital Lille
(Lille, France), and a specialized care home (France).
Experiments were performed under the supervision of their treating physician.
Participants were recruited based on the following criteria.
To qualify for inclusion, participants must:
\begin{enumerate}
	\item be at least 18 years old and no older than 60
	years,
  \item belong to class 2 or 3 according to the \ac{bci}	user selection criteria
    presented by~\textcite{Wolpaw2006},\label{item:patients/inclusion/wolpaw}
  \item have limitations to the extent or comfort of their eye motor control\label{item:patients/inclusion/oculomotor}
\end{enumerate}
Participants were excluded if they:
\begin{enumerate}
  \item had a diagnosis of a major medical condition, including any major
    neurological or psychiatric disorder other than those of interest based on
    inclusion criteria~\ref{item:patients/inclusion/wolpaw},
    and~\ref{item:patients/inclusion/oculomotor}\label{item:patients/exclusion/medical}
  \item had a predisposition to or a history of any kind of epileptic seizures,
    including photosensitive epilepsy,\label{item:patients/exclusion/epilepsy}
  \item had a severe loss in vision or hearing that would significantly impair
        participation in the experiment,\label{item:patients/exclusion/vision}
  \item are currently using specific psychoactive medications or substances that could affect the outcome.\label{item:patients/exclusion/cognitive}
  \item were unable to understand the experiment instructions and cooperate,
  \item had any other limitations preventing them from performing the given task.
\end{enumerate}

@@

Another limiting factor is the difficulty in correctly interpreting eye
tracker results in studies with individuals with gaze impairments.
If eye tracking is possible at all, it is not guaranteed that the user is able
to successfully perform the calibration procedure.
Further experiments should be carried out with a stationary eye tracker with
more advanced capabilities, although systems using a head fixator or headrest
should be avoided.
This is not practical when working with
wheelchair-bound individuals who might have undergone a tracheostomy and may
suffer from spasticity.

>>>>

\Ac{bci} assistive technologies for communication~\cite{Millan2010}
target individuals with \acf{sspi}~\cite{Peters2022}.
Visual \acp{bci}, which rely on the interpretation of visual stimuli by the user,
offer several advantages in this context.
They can work with non-invasive recording technology and can use rapid
stimulation.
This makes them well-suited for real-time communication tasks.

@@

Several studies with visual oddball \acp{bci} show that performance drops when not fixating the intended
target~\cite{Brunner2010, Treder2010, RonAngevin2019}, necessitating
gaze-in\-de\-pen\-dent solutions.
These studies build on the assumption that \ac{bci} users with \ac{sspgi}
would feel comfortable operating an interface in pure covert \ac{vsa} with
central fixation.
One could argue that a \ac{bci} that is only verified to work when central
fixation is maintained could also be considered gaze-dependent.
This does not account for the residual eye motor capabilities of most people
with \ac{sspi}, the (dis)comfort they experience while
performing gaze fixation and other confounding factors resulting from their eye
motility.

@@

\textcite{Lesenfants2014} tested a \ac{bci} using gaze-independent \ac{ssvep} in six
participants with \ac{lis} yet only exceeded chance level accuracy in two.
More recently, \textcite{Peters2020} performed a trial with two
participants with late-stage \ac{als} and visual impairment.
Their \ac{ssvep} interface was not optimized for gaze-independence, but the
system showed high accuracy, outperforming an eye tracking alternative.
It would be of interest to verify if such results can be replicated with
participants with other conditions, and with a visual oddball \acp{bci}.

@@

\section{Materials \& methods}
\subsection{Recruitment}
Participants were recruited across the Neuromuscular Reference center at
University Hospital Leuven (Leuven, Belgium), TRAINM Neuro Rehab Clinics
(Antwerp, Belgium), the Neurorehabilitation Unit at University Hospital Lille
(Lille, France), and a specialized care home (France).
Experiments were performed under the supervision of their treating physician.
Participants were recruited based on the following criteria.
To qualify for inclusion, participants must:
\begin{enumerate}
	\item be at least 18 years old and no older than 60
	years,
  \item belong to class 2 or 3 according to the \ac{bci}	user selection criteria
    presented by~\textcite{Wolpaw2006},\label{item:patients/inclusion/wolpaw}
  \item have limitations to the extent or comfort of their eye motor control\label{item:patients/inclusion/oculomotor}
\end{enumerate}
Participants were excluded if they:
\begin{enumerate}
  \item had a diagnosis of a major medical condition, including any major
    neurological or psychiatric disorder other than those of interest based on
    inclusion criteria~\ref{item:patients/inclusion/wolpaw},
    and~\ref{item:patients/inclusion/oculomotor}\label{item:patients/exclusion/medical}
  \item had a predisposition to or a history of any kind of epileptic seizures,
    including photosensitive epilepsy,\label{item:patients/exclusion/epilepsy}
  \item had a severe loss in vision or hearing that would significantly impair
        participation in the experiment,\label{item:patients/exclusion/vision}
  \item are currently using specific psychoactive medications or substances that could affect the outcome.\label{item:patients/exclusion/cognitive}
  \item were unable to understand the experiment instructions and cooperate,
  \item had any other limitations preventing them from performing the given task.
\end{enumerate}

@@

Another limiting factor is the difficulty experienced in correctly interpreting eye
tracker results in studies with individuals with gaze impairments.
If eye tracking is possible at all, it is not guaranteed that the user is able
to successfully perform the calibration procedure.
Further experiments should be carried out with a stationary eye tracker with
more advanced capabilities, although systems using a head fixator or headrest
should be avoided.
This is not practical when working with
wheelchair-bound individuals who might have undergone a tracheostomy and may
suffer from spasticity.
